Sensor Clusterization in D-optimal Design in Infinite Dimensional Bayesian Inverse Problems

### 2020-07-23 paper.tex
Submitted to arxivgit add paper.tex git add paper.tex git add paper.tex git add paper.tex git add paper.tex git add paper.tex git add paper.tex git add paper.tex !

\documentclass{amsart}
\input{definitions}

@@

\author{Yair Daon}
\address{Courant Institute}
%% \curraddr{}
%% \email{yair.daon@gmail.com}
%% \thanks{}

%    The 2010 edition of the Mathematics Subject Classification is
%    the current definitive version.
\subjclass{Primary: 
  62F15, % Statistics - bayesian inference
  35R30, % PDE - Inverse problems
  65C50; % Numerical analysis - other computational problems in probability
  Secondary:
  28C20, % Measure and integration - set functions and measures and integrals in
  %infinite-dimensional spaces (Wiener measure, Gaussian measure, etc.
}

\date{\today}

\begin{abstract}
  We investigate the problem of sensor clusterization in optimal
  experimental design for infinite-dimensional Bayesian inverse
  problems. We suggest an analytically tractable model for such
  designs and reason how it may lead to sensor clusterization in the
  case of iid measurement noise. We also show that in the case of
  spatially correlated measurement error clusterization does not
  occur. Our analysis suggests a novel method for finding optimal
  designs. As a part of the analysis we prove a matrix determinant
  lemma analog in infinite dimensions, as well as a lemma for
  calculating derivatives of $\log \det$ of operators.
\end{abstract}

@@

Any of the above mentioned approaches might serve as a remedy and push
sensors away from each other. Yet, none offers any insight as to why
clusterization occurs. Also, as better models are employed, the model
error decreases and the clusterization phenomenon will reappear. While
these approaches are practical and help us avoid the problem, they do
not provide insight as to why sensors are clustering.

\subsection{Contribution}
We propose and thoroughly study an analytically tractable model for
understanding D-optimal designs. In this model, an optimal design
arises as a solution to a constrained optimization problem. This
problem is solved using Lagrange multipliers (section \ref{section:D
  and grad}). In the case of no model error we show this problem is in
fact a set of eigenvector-eigenvalue problems (equation
\eqref{eq:eigenproblem}). We then characterize an optimal design as
one that reduces the highest uncertainties, and show how that can give
rise to repeated measurements (figure \ref{fig:clusterization}).

The solution to the this could be used to seek a design in the real
problem we are interested in and we outline this in section
\ref{section:conclusion}. The analysis that culminates in equation
\eqref{eq:true target} also gives us an upper bound for the design
criterion (denoted $\tar$) below.

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings --- the Matrix Determinant Lemma and a
lemma for calculating the derivative of determinants. We also provide
other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemmas \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}. We also show that in our model with independent measurement
noise, an interesting phenomenon arises, in which the eigenvectors of
the measurement operator (denoted below $\obs^*\obs$) are the same as
the eigenvectors of the prior covariance in observation space, which
helps arrive at the results outlined in the previous paragraph. The
arguments presented are completely generic and apply equally to any of
the scenarios mentioned above.

@@

\subsection{An Example of Clusterization}\label{subsec:example}
\begin{figure}
  \begin{tikzpicture}[thick, scale=1.3, every node/.style={scale=0.99}]
    \begin{axis}
      [
      xmin = 0,
      xmax = 3.14,
      xlabel = {$x$},
      ylabel = posterior std,
      ymin   = 0,
      %compat = 1.3,
      % ymax   = 130,
      % ytick = \empty,
      legend cell align=left,
      % legend style={at={(0.45,0.2)}}
      legend pos= outer north east 
      ]
      % \draw[black!30!white, thin] (50,0) -- (50,130);
      % 
      \addplot [thin, black, mark=none] table{data/stdv-heat-sens1-var1.txt};
      \addlegendentry{1 sensors};

      \addplot [thin, blue, mark=none] table{data/stdv-heat-sens2-var1.txt};
      \addlegendentry{2 sensors};

      \addplot [thin, red, mark=none] table{data/stdv-heat-sens3-var1.txt};
      \addlegendentry{3 sensors};

      \addplot [thin, green, mark=none] table{data/stdv-heat-sens4-var1.txt};
      \addlegendentry{4 sensors};

      \addplot [thin, purple, mark=none] table{data/stdv-heat-sens5-var1.txt};
      \addlegendentry{5 sensors};

      \addplot [thin, cyan, mark=none] table{data/stdv-heat-sens6-var1.txt};
      \addlegendentry{6 sensors};

      \addplot [black,  only marks, mark=x, mark size=1.5] 
      table{data/locs-heat-sens1-var1.txt}; 
      \addplot [blue,   only marks, mark=x, mark size=1.5]
      table{data/locs-heat-sens2-var1.txt}; 
      \addplot [red,    only marks, mark=x, mark size=1.5]
      table{data/locs-heat-sens3-var1.txt};
      \addplot [green,  only marks, mark=x, mark size=1.5] 
      table{data/locs-heat-sens4-var1.txt}; 
      \addplot [purple, only marks, mark=x, mark size=1.5] 
      table{data/locs-heat-sens5-var1.txt}; 
      \addplot [cyan,   only marks, mark=x, mark size=1.5] 
      table{data/locs-heat-sens6-var1.txt};

    \end{axis}
  \end{tikzpicture}
  \caption{Shown are pointwise posterior standard deviations for
    various numbers of sensors, and location of these sensors for the
    1D heat equation with homogeneous Dirichlet boundary described in
    section \ref{subsec:example}. Note that the observation data does
    not influence the pointwise variances, so there is no point in
    showing the posterior mean.}
  \label{fig:clusterization illustration}
\end{figure}
In section \ref{section:abstract OED} we present a more abstract and
general formulation of the problem, but for the purpose of
illustration, we present the problem via a toy model --- the 1D heat
equation in $[0,\pi]$ with homogeneous Dirichlet boundary
conditions.

@@

We would like to infer the initial condition $u_0$. For that purpose,
we measure $u$ at some set of locations $\x_j \in [0,\pi], j=1,...,m$
and a final time $T > 0$. We assume centered Gaussian measurement
error, so we can observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2), \sigma > 0$ iid.  We model the
initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with homogeneous Dirichlet boundary condition. It is
well known \cite{Tarantola05} that for linear problems, with Gaussian
prior and error, the posterior is also Gaussian with a covariance that
does not depend on the observed data. The posterior covariance
$\postcov$ is known to have a closed form formula, even in infinite
dimensions\cite{Stuart10}. If we denote by $\fwd$ the dynamics
operator, so that $u( \cdot,T) = \fwd u_0$, and the observation
operator $\obs$ so that $u(\x_j,T) = (\obs u)_j, j=1,...,m$, then the
posterior covariance is known and depends on $\prcov, \fwd, \obs$ and
$\sigma^2$ (see section \ref{section:abstract OED} and
\eqref{eq:postcov} specifically).

We will consider the information-theoretic design criterion presented
in section \ref{section:OED intro}. Thus, we choose $\x_j,j=1,...,m$
so as to minimize an expression analogous to the log-determinant of
the posterior covariance operator (corresponding to a D-optimal
design). We can see the clusterization effect in figure
\ref{fig:clusterization illustration} --- the design criterion
suggests placing sensors in locations that are extremely close.

\subsection{Remarks to the Reader}\label{subsec:plan}
The Lagrange multipliers problem for iid measurement noise is
discussed in section \ref{section:vanishing}. In that section, it is
shown how sensor clusterization occurs. It is advised that a reader
less comfortable with Hilbert and Banach spaces mostly ignore these
terms and consider $\hil$ and $\ban$ finite dimensional vector spaces.

%% Seeking an optimal design amounts to maximizing $\tar(\obs)$ in
%% $\obs$, subject to constraints posed on $\obs$. These enter as
%% constraints on the allowed set of measurement vectors we can choose
%% from --- the rows of $\obs$. These constraints are determined by the
%% problem structure --- for example, in the 1D heat equation example
%% considered in section \ref{subsec:example}, we were constrained to
%% choose $\obs = (\meas_1,...,\meas_m)^t$ with $\meas_j = \delta_{\x_j}$
%% for some $\x_j \in [0,\pi]$. We present an approach to understanding
%% optimal experiment design based on the first-order necessary condition
%% for optimal design, where the constraints are incorporated using
%% Lagrange multipliers. This condition is used to understand what causes
%% clusterization. One problem with such an approach is that the set we
%% can choose $\obs$ from is complicated and difficult to work with. For
%% example, if $\ban = C(\Omega)$ for some compact $\Omega$, then the
%% allowed set of measurement vectors may only be point evaluations --- a
%% set that is not easy to describe as a level set of some simple
%% function, a description crucial for the use of Lagrange multipliers.
%% In order to facilitate analysis, we consider an easier constraint in
%% section \ref{subsec:unit length}. Reverting to this simpler constraint
%% is a limitation of our analysis but it provides some insight into the
%% cause of clusterization.

@@

The theoretical foundations for inverse problems over function spaces
can be found in \cite{Stuart10}. At this point, it is useful to
record a known result regarding the posterior covariance operator of
our inverse problem, namely
\begin{align}\label{eq:postcov precursor}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd )^{-1}.
\end{align}

@@

We want to infer $\param \in \hil$, some parameter of the dynamics
given noisy observations of $\fwd \param$. We model $\param \sim
\normal(0,\prcov)$ with some appropriate covariance operator $\prcov$
\cite{Stuart10}. Observations (measurements) are taken via the linear
observation operator $\obs \in ( \ban^* )^m$, where $m$ is the number
of observations allowed (e.g. sensors at our disposal). In an analogy
with linear algebra (where row vectors are thought of as linear
functionals), we think of $\obs$ as having ``rows'' $\meas_j,
j=1,...,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,..., \meas_m)^t, \meas_j \in \ban^*, j = 1,...,m.
\end{equation}
This way, for $u \in \ban$ we have $\obs u = (\meas_1(u), ...,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
A few observations regarding $\obs$ are in order. First, it is good to
keep in mind that $( \ban^* )^m$ is a Banach space with norm $\| \obs
\| = \sum_{j=1}^m \|\meas_j\|$. Next, for $u \in \ban$ and $v\in
\R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

\begin{observation}
  It is best to think of $\meas_j,j=1,...,m$ as row vectors and of
  $\obs$ as a matrix with rows $\meas_j$. 
\end{observation}

@@

Usually, this is written using $\tmp := \obs \fwd$ and $\tmp^*$ but
since we need to separate the observation operator $\obs$ from the
forward operator $\fwd$ we write the explicit expression
\eqref{eq:postcov}. Taking $\modcov = 0$ is common practice
\cite{Tarantola05,KaipioSomersalo05,Vogel02} and then we are reduced
to the case where we take iid observations and $\Sigma = \sigma^2I$ is
simply a scalar matrix that does not depend on $\obs$. Note that
taking an error model with a non-scalar covariance as we do here
allows us to consider model error (modeled by $\modcov$) as well as
measurement error (modeled by $\sigma^2$). For example, say we believe
our forward model does not capture some small scale phenomenon.  Then
we may express this belief by saying $\tru = \fwd + \err$, with $\fwd$
depending on $\param$ and $\err$ depending on $\sspar$, and $\param
\perp \sspar$. We do not know much about this effect but it is
reasonable to assume it changes continuously in our domain. We (may
choose to) model it as $\normal (0, \modcov)$ and take $\modcov$ to
reflect the spatial (or other) variability we imagine $\err \sspar$
has. Such small scale phenomenon can arise as a modeling issue, where
we might not model the system in its entirety. It can also arise from
a numerical source, where our discretization of the system is not fine
enough to capture all small scale phenomena. We will see that assuming
some correlation in the error may reduce the clusterization
phenomenon, as is also reported in the literature \cite{Ucinski05}.

\subsection{D-Optimal Design in Infinite Dimensions} 
As pointed out before, we consider D-optimal experimental design. The
meaning of this notion in infinite-dimensional Hilbert spaces was
investigated in \cite{AlexanderianGloorGhattas14}. Note that the
authors make assumptions that amount to $\Sigma=I$ (implied by
$\modcov = 0,\sigma^2=1$), but we choose not to take these
simplification here. This is because $\Sigma$ can determine ``how
much'' clusterization we see. As stated
\cite[pp. 681]{AlexanderianGloorGhattas14}, the results hold for more
general covariance matrices. The derivation of the design criterion
from information theoretic considerations can be verified similarly.
The conclusion is that in infinite-dimensions, a D-optimal design is
well-defined as maximizing the expected KL divergence between
posterior and prior and is summarized, using our notation, in the
following theorem:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]
  Let $\pr = \normal(0,\prcov)$ be a centered Gaussian prior
  on $\hil$   and let $\post = \normal(\postmean,\postcov)$ 
  the posterior measure on $\ban$ for the Bayesian linear
  inverse problem $\data =  \obs \fwd\param + \obs \eps' + \eps$ discussed
  above. Then 
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

\section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D and grad}
From \eqref{eq:objective} we wish to characterize solution(s) of the
following optimization problem for $\tar$. %%: (\ban^*)^m \to \R$:
\begin{align*}
  \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
  = \argmax_{\obs} \frac12 \log \det 
  (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
\end{align*}
where $\obs$ is constrained to some allowed set of measurements
(e.g.\ pointwise evaluations on $[0,\pi]$)
%%, for the example in section \ref{subsec:example}).
We seek a formulation of the optimal design using Lagrange
multipliers. We first find gradients in function spaces.

@@

As mentioned before, we cannot choose any $\obs$ when maximizing
$\tar(\obs)$. Recall $\obs = (\meas_1,...,\meas_m)^t \in (\ban^*)^m$
and each $\meas_j,j=1,...,m$ must be chosen from some allowed set of
functionals in $\ban^*$. This set differs based on the kind of sensors
we have at our disposal and the properties of $\ban$, as discussed in
section \ref{subsec:dynamics}. The following proposition will give us
better understanding of the constraints. 
\begin{proposition}\label{prop:bigger better}
  Let $\obs = (\meas_1,...,\meas_m)^t$, $j \in \{1,...,m\}$, $\sigma^2
  > 0$ and $\lambda > 1$. Then $\tar(\obs)$ increases if we use
  $\lambda \meas_j$ in $\obs$ instead of $\meas_j$.
\end{proposition}
Proof of Proposition \ref{prop:bigger better} is delegated to the
appendix but the idea behind it is simple: making rows of $\obs$ large
is equivalent to making $\sigma^2$ small. This can be understood
easily from the formulation of the inverse problem $\data = \obs \fwd
\param + \obs \eps' + \eps$, where $\eps'$ can be taken as model error
and $\eps$ is iid measurement error.

>>>>

\documentclass{amsart}

\input{definitions}

@@

\author{Yair Daon}
\address{Porter School of the Environment and Earth
  Sciences, Tel Aviv University\\ Tel Aviv, Israel}
%% \curraddr{Porter School of Environment Studies, Tel Aviv University\\ Tel Aviv, Israel}
%% \curraddr{}
\email{yair.daon@gmail.com}
%% \thanks{}

%    The 2010 edition of the Mathematics Subject Classification is
%    the current definitive version.
\subjclass{Primary: 
  62F15, % Statistics - bayesian inference
  35R30, % PDE - Inverse problems
  Secondary:
  28C20, % Measure and integration - set functions and measures and integrals in
  %infinite-dimensional spaces (Wiener measure, Gaussian measure, etc.
}
%% Primary: 62F15, 35R30; Secondary: 28C20
\date{\today}

\begin{abstract}
  We investigate the problem of sensor clusterization in optimal
  experimental design for infinite-dimensional Bayesian inverse
  problems. We suggest an analytically tractable model for such
  designs and reason how it may lead to sensor clusterization in the
  case of iid measurement noise. We also show that in the case of
  spatially correlated measurement error clusterization does not
  occur. As a part of the analysis we prove a matrix determinant
  lemma analog in infinite dimensions, as well as a lemma for
  calculating derivatives of $\log \det$ of operators.
\end{abstract}

@@

Any of the above mentioned approaches might serve as a remedy and push
sensors away from each other. Yet, none offers any insight as to why
clusterization occurs. Also, as better models are employed, model
error is decreased and the clusterization phenomenon will eventually
reappear. While these approaches are practical and help us avoid the
problem, they do not provide insight as to why sensors are clustering.

\subsection{Contribution}
We propose and thoroughly study an analytically tractable model for
understanding D-optimal designs. In this model, an optimal design
arises as a solution to a constrained optimization problem, formulated
using Lagrange multipliers (section \ref{section:D and grad}). In the
case of no model error we show this problem is in fact a set of
eigenvector-eigenvalue problems (equation \eqref{eq:eigenproblem}). We
then characterize an optimal design as one that reduces the highest
uncertainties, and show how that can give rise to repeated
measurements (section \ref{subsec:clusterization} and figure
\ref{fig:clusterization}).

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings --- the Matrix Determinant Lemma and a
lemma for calculating the derivative of determinants. We also provide
other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemmas \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}. %%  We also show that in our model with independent measurement
%% noise, an interesting phenomenon arises, in which the eigenvectors
%% of the measurement operator (denoted below $\obs^*\obs$) are the
%% same as the eigenvectors of the prior covariance in observation
%% space, which helps arrive at the results outlined in the previous
%% paragraph.
The arguments presented are completely generic and apply equally to
any of the scenarios mentioned in the introduction.

@@

\subsection{An Example of Clusterization}\label{subsec:example}
\begin{figure}
  \begin{tikzpicture}[thick, scale=1.3, every node/.style={scale=0.99}]
    \begin{axis}
      [
      xmin = 0,
      xmax = 3.14,
      xlabel = {$x$},
      ylabel = posterior std,
      ymin   = 0,
      %compat = 1.3,
      % ymax   = 130,
      % ytick = \empty,
      legend cell align=left,
      % legend style={at={(0.45,0.2)}}
      legend pos= outer north east 
      ]
      % \draw[black!30!white, thin] (50,0) -- (50,130);
      % 
      %% \addplot [thin, black, mark=none] table{stdv-heat-sens1-var1.txt};
      %% \addlegendentry{1 sensors};

      %% \addplot [thin, blue, mark=none] table{stdv-heat-sens2-var1.txt};
      %% \addlegendentry{2 sensors};

      %% \addplot [thin, red, mark=none] table{stdv-heat-sens3-var1.txt};
      %% \addlegendentry{3 sensors};

      \addplot [thin, green, mark=none] table{stdv-heat-sens4-var1.txt};
      \addlegendentry{4 sensors};

      \addplot [thin, purple, mark=none] table{stdv-heat-sens5-var1.txt};
      \addlegendentry{5 sensors};

      \addplot [thin, cyan, mark=none] table{stdv-heat-sens6-var1.txt};
      \addlegendentry{6 sensors};

      %% \addplot [black,  only marks, mark=x, mark size=1.5] 
      %% table{locs-heat-sens1-var1.txt}; 
      %% \addplot [blue,   only marks, mark=x, mark size=1.5]
      %% table{locs-heat-sens2-var1.txt}; 
      %% \addplot [red,    only marks, mark=x, mark size=1.5]
      %% table{locs-heat-sens3-var1.txt};
      \addplot [green,  only marks, mark=x, mark size=1.5] 
      table{locs-heat-sens4-var1.txt}; 
      \addplot [purple, only marks, mark=x, mark size=1.5] 
      table{locs-heat-sens5-var1.txt}; 
      \addplot [cyan,   only marks, mark=x, mark size=1.5] 
      table{locs-heat-sens6-var1.txt};

    \end{axis}
  \end{tikzpicture}
  \caption{Shown are pointwise posterior standard deviations for
    various numbers of sensors, and location of these sensors for the
    1D heat equation with homogeneous Dirichlet boundary described in
    section \ref{subsec:example}. Note the clusterization that occurs
    when 6 sensors are used --- only 4 measurement locations are
    included. Observed data does not influence the pointwise
    variances, hence it is omitted.}
\end{figure}\label{fig:clusterization illustration}
In section \ref{section:abstract OED} we present a more abstract and
general formulation of the problem, but for the purpose of
illustration, we present the problem via a toy model --- the 1D heat
equation in $[0,\pi]$ with homogeneous Dirichlet boundary
conditions.

@@

We would like to infer the initial condition $u_0$. For that purpose,
we measure $u$ at some set of locations $\x_j \in [0,\pi], j=1, \dots,m$
and a final time $T > 0$. We assume centered Gaussian measurement
error, so we can observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2), \sigma > 0$ iid.  We model the
initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with homogeneous Dirichlet boundary condition. It is
well known \cite{Tarantola05} that for linear problems, with Gaussian
prior and error, the posterior is also Gaussian with a covariance that
does not depend on the observed data. The posterior covariance
$\postcov$ is known to have a closed form formula, even in infinite
dimensions\cite{Stuart10}. If we denote by $\fwd$ the dynamics
operator, so that $u( \cdot,T) = \fwd u_0$, and the observation
operator $\obs$ so that $u(\x_j,T) = (\obs u)_j, j=1,\dots,m$, then the
posterior covariance is known and depends on $\prcov, \fwd, \obs$ and
$\sigma^2$ (see section \ref{section:abstract OED} and
\eqref{eq:postcov} specifically).

We will consider generalization of the information-theoretic design
criterion presented in section \ref{section:OED intro} to infinite
dimensions (section \ref{subsec:D optimal design}). We choose
$\x_j,j=1,\dots,m$ so as to minimize an expression analogous to the
log-determinant of the posterior covariance operator (corresponding to
a D-optimal design). We can see the clusterization effect in figure
\ref{fig:clusterization illustration} --- the design criterion
suggests placing sensors in locations that are extremely close.

\subsection{Remarks to the Reader}\label{subsec:plan}
The Lagrange multipliers problem for iid measurement noise is
discussed in section \ref{section:vanishing}. In that section, it is
shown how sensor clusterization can occur. The reader may skip if they
are constrained by time. It is advised that a reader less comfortable
with Hilbert and Banach spaces mostly ignore these terms and consider
$\hil$ and $\ban$ finite dimensional vector spaces.

%% Seeking an optimal design amounts to maximizing $\tar(\obs)$ in
%% $\obs$, subject to constraints posed on $\obs$. These enter as
%% constraints on the allowed set of measurement vectors we can choose
%% from --- the rows of $\obs$. These constraints are determined by the
%% problem structure --- for example, in the 1D heat equation example
%% considered in section \ref{subsec:example}, we were constrained to
%% choose $\obs = (\meas_1,\dots,\meas_m)^t$ with $\meas_j = \delta_{\x_j}$
%% for some $\x_j \in [0,\pi]$. We present an approach to understanding
%% optimal experiment design based on the first-order necessary condition
%% for optimal design, where the constraints are incorporated using
%% Lagrange multipliers. This condition is used to understand what causes
%% clusterization. One problem with such an approach is that the set we
%% can choose $\obs$ from is complicated and difficult to work with. For
%% example, if $\ban = C(\Omega)$ for some compact $\Omega$, then the
%% allowed set of measurement vectors may only be point evaluations --- a
%% set that is not easy to describe as a level set of some simple
%% function, a description crucial for the use of Lagrange multipliers.
%% In order to facilitate analysis, we consider an easier constraint in
%% section \ref{subsec:unit length}. Reverting to this simpler constraint
%% is a limitation of our analysis but it provides some insight into the
%% cause of clusterization.

@@

The theoretical foundations for inverse problems over function spaces
can be found in \cite{Stuart10}. This section is mostly devoted to
setting notation. At this point, it is useful to record a known result
regarding the posterior covariance operator of our inverse problem,
namely
\begin{align}\label{eq:postcov precursor}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd )^{-1}.
\end{align}

@@

We want to infer $\param \in \hil$, some parameter of the dynamics
given noisy observations of $\fwd \param$. We model $\param \sim
\normal(0,\prcov)$ with some appropriate covariance operator $\prcov$
\cite{Stuart10}. Observations (measurements) are taken via the linear
observation operator $\obs \in ( \ban^* )^m$, where $m$ is the number
of observations allowed (e.g. sensors at our disposal). In an analogy
with linear algebra (where row vectors are thought of as linear
functionals), we think of $\obs$ as having ``rows'' $\meas_j,
j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \ban^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \ban$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
A few observations regarding $\obs$ are in order. First, it is good to
keep in mind that $( \ban^* )^m$ is a Banach space with norm $\| \obs
\| = \sum_{j=1}^m \|\meas_j\|$. Next, for $u \in \ban$ and $v\in
\R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

\begin{observation}
  It is best to think of $\meas_j,j=1,\dots,m$ as row vectors and of
  $\obs$ as a matrix with rows $\meas_j$. 
\end{observation}

@@

%% Usually, this is written using $\tmp := \obs \fwd$ and $\tmp^*$ but
%% since we need to separate the observation operator $\obs$ from the
%% forward operator $\fwd$ we write the explicit expression
%% \eqref{eq:postcov}.
Taking $\modcov = 0$ is common practice
\cite{Tarantola05,KaipioSomersalo05,Vogel02} and then we are reduced
to the case where we take iid observations and $\Sigma = \sigma^2I$ is
simply a scalar matrix that does not depend on $\obs$. Note that
taking an error model with a non-scalar covariance as we do here
allows us to consider model error (modeled by $\modcov$) as well as
measurement error (modeled by $\sigma^2$). For example, say we believe
our forward model does not capture some small scale phenomenon.  Then
we may express this belief by saying $\tru = \fwd + \err$, with $\fwd$
depending on $\param$ and $\err$ depending on $\sspar$, and $\param
\perp \sspar$. We do not know much about this effect but it is
reasonable to assume it changes continuously in our domain. We (may
choose to) model it as $\normal (0, \modcov)$ and take $\modcov$ to
reflect the spatial (or other) variability we imagine $\err \sspar$
has. Such small scale phenomenon can arise as a modeling issue, where
we might not model the system in its entirety. It can also arise from
a numerical source, where our discretization of the system is not fine
enough to capture all small scale phenomena. We will see that assuming
some correlation in the error may reduce the clusterization
phenomenon, as is also reported in the literature \cite{Ucinski05}.

\subsection{D-Optimal Design in Infinite Dimensions}\label{subsec:D optimal design} 
The meaning of D-optimal design in infinite-dimensional Hilbert spaces
was investigated in \cite{AlexanderianGloorGhattas14}. Note that the
authors make assumptions that amount to $\Sigma=I$ (implied by
$\modcov = 0,\sigma^2=1$), but we choose not to take these
simplification here. This is because $\Sigma$ can determine ``how
much'' clusterization we see. As stated
\cite[pp. 681]{AlexanderianGloorGhattas14}, the results hold for more
general covariance matrices. The conclusion is that in
infinite-dimensions, a D-optimal design is well-defined as maximizing
the expected KL divergence between posterior and prior and is
summarized, using our notation, in the following theorem:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]
  Let $\pr = \normal(0,\prcov)$ be a centered Gaussian prior
  on $\hil$   and let $\post = \normal(\postmean,\postcov)$ 
  the posterior measure on $\ban$ for the Bayesian linear
  inverse problem $\data =  \obs \fwd\param + \obs \eps' + \eps$ discussed
  above. Then 
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

\section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D and grad}
From \eqref{eq:objective} we wish to characterize solution(s) of the
following optimization problem for $\tar$. %%: (\ban^*)^m \to \R$:
\begin{align}\label{eq:optimization}
  \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
  = \argmax_{\obs} \frac12 \log \det 
  (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
\end{align}
where $\obs$ is constrained to some allowed set of measurements
(e.g.\ pointwise evaluations on $[0,\pi]$)
%%, for the example in section \ref{subsec:example}).
We seek a formulation of the optimal design using Lagrange
multipliers. In section \ref{section:objective} we find the gradient
of $\tar$.  In section \ref{subsec:unit length} we relax the
constraints on $\obs$ so that the optimization problem in
\eqref{eq:optimization} and find gradients for the new constraints.

@@

As mentioned before, we cannot choose any $\obs$ when maximizing
$\tar(\obs)$. Recall $\obs = (\meas_1,\dots,\meas_m)^t \in (\ban^*)^m$
and each $\meas_j,j=1,\dots,m$ must be chosen from some allowed set of
functionals in $\ban^*$. This set differs based on the kind of sensors
we have at our disposal and the properties of $\ban$, as discussed in
section \ref{subsec:dynamics}. The following proposition will give us
better understanding of the constraints. 
\begin{proposition}\label{prop:bigger better}
  Let $\obs = (\meas_1,\dots,\meas_m)^t$, $j \in \{1,\dots,m\}$, $\sigma^2
  > 0$ and $\lambda > 1$. Then $\tar(\obs)$ increases if we use
  $\lambda \meas_j$ in $\obs$ instead of $\meas_j$.
\end{proposition}
Proof of Proposition \ref{prop:bigger better} is delegated to the
appendix but the idea behind it is simple: making rows of $\obs$ large
is equivalent to making $\sigma^2$ small. This can be understood
easily from the formulation of the inverse problem $\data = \obs \fwd
\param + \obs \eps' + \eps$, where $\eps'$ can be taken as model error
and $\eps$ is iid measurement error.
